---
title: "Standard protocols, DNN training and QuPath/StarDist exchange"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Standard protocols, DNN training and QuPath/StarDist exchange}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(segmantR)
```

This vignette walks through the reproducible, exchangeable side of
segmantR: versioned segmentation protocols, the neutral interchange format,
training datasets and model bundles, and the QuPath/StarDist adapter.
Everything executed here runs in R alone; steps that need Python or QuPath
are shown but not run.

## 1. Standard protocols

Protocols are declarative JSON definitions with whitelisted parameters.

```{r}
protocols <- sg_protocol_list()
protocols[, c("id", "version", "status", "backend")]
```

Each parameter has a type, bounds, unit and calibration requirement:

```{r}
spec <- sg_protocol_schema("threshold.otsu.v1")
spec[!spec$name %in% c("band_a_nm", "band_b_nm", "band_min_nm",
                       "band_max_nm"),
     c("name", "type", "unit", "maps_to")]
```

Running a protocol delegates to the direct function with a recorded argument
mapping. The run envelope makes declarative and direct runs comparable:

```{r}
img <- sg_example_image("fluorescence_nuclei")
run <- sg_protocol_run(img, "threshold.otsu.v1", min_area = 5L,
                       output = "run")
run
str(run$record$delegate$arguments)
direct <- sg_segment_threshold(img, method = "otsu", min_area = 5L)
identical(direct$labels, run$mask$labels)
```

Invalid input is refused with classified conditions instead of being
guessed:

```{r}
res <- sg_protocol_validate("threshold.otsu.v1", img, min_area = -1L,
                            error = FALSE)
res$error$code
try(sg_protocol_run(img, "mesmer.2d.v1"))
```

Post-processing is a protocol too:

```{r}
clean <- sg_protocol_run(run$mask, "postprocess.label-cleanup.v1",
                         min_area = 40L, fill_holes = TRUE)
clean$n_cells
```

### Hyperspectral channels

Bands are selected by index, name, wavelength or a registered operation;
value semantics are declared, never inferred.

```{r}
cube <- array(runif(32 * 32 * 3), dim = c(32, 32, 3))
hsi <- new_sg_image(cube, channels = c("b460", "b550", "b650"),
                    bands = data.frame(wavelength_nm = c(460, 550, 650)),
                    value_semantics = "reflectance")
nd <- sg_select_channel(hsi, band_operation = "normalized_difference",
                        band_a_nm = 650, band_b_nm = 550)
nd$provenance$band_selection[c("operation", "wavelength_nm")]
```

Optional DNN protocols check their backend when they run and never fall
back to another method:

```{r}
dnn <- sg_protocol_run(img, "stardist.2d.v1", output = "run")
dnn$record$status
```

## 2. Neutral interchange

A bundle contains an integer label TIFF, exact pixel-edge GeoJSON with
stable object ids, the legend, long-format measurements, a manifest and a
SHA-256 inventory.

```{r}
dest <- file.path(tempdir(), "segmantR-bundle")
unlink(dest, recursive = TRUE)
bundle <- sg_export_interchange(run, dest, image = img)
list.files(dest)
bundle$bundle_digest
```

Imports verify integrity, schema, dtype, orientation, plane, legend and the
agreement between GeoJSON and TIFF, and return a **staged** mask:

```{r}
imported <- sg_import_interchange(dest,
                                  expected_digest = bundle$bundle_digest)
imported$checks[, c("check", "status")]
sg_mask_status(imported$mask)$status
```

Reviewed masks are protected:

```{r}
reviewed <- sg_review_mask(imported$mask,
                           expected_revision = sg_mask_revision(imported$mask))
candidate <- clean
try(sg_replace_mask(reviewed, candidate))
replaced <- sg_replace_mask(reviewed, candidate,
                            expected_revision = sg_mask_revision(reviewed),
                            overwrite = TRUE)
sg_mask_status(replaced)$status
```

## 3. Training data, StarDist training and prediction

Dataset manifests tile image/mask pairs and split them by origin so tiles of
one subject never appear in two splits.

```{r}
mask <- sg_cleanup_labels(sg_example_mask("fluorescence_nuclei"),
                          disconnected = "split")
pairs <- lapply(c("A", "B", "C"), function(s) {
  list(image = img, mask = mask, subject_id = s)
})
ts <- sg_prepare_training_data(pairs, tile_size = 32L,
                               split = c(train = 0.34, validation = 0.33,
                                         test = 0.33))
ts
sg_validate_training_manifest(ts$manifest, mask_type = "instance")$ok
```

A foreground/background mask is not accepted as instance training data:

```{r}
binary <- new_sg_mask(ifelse(mask$labels > 0L, 1L, 0L))
try(sg_dataset_manifest(list(list(image = img, mask = binary,
                                  subject_id = "A")), tile_size = 32L))
```

Training needs the Python `stardist` and `tensorflow` modules; all data
checks run before Python is touched:

```{r, eval = FALSE}
model <- sg_train_stardist(ts, n_epochs = 50L, save_path = "models",
                           model_card = list(
                             data_domain = "synthetic fluorescence nuclei",
                             limitations = "demonstration only",
                             license = list(id = "CC-BY-4.0")))
sg_package_model(model, "nuclei.segmantR")
loaded <- sg_load_model("nuclei.segmantR")   # checksums verified
pred <- sg_predict_model(loaded, img, output = "run")
sg_mask_status(pred$mask)$status              # "staged"
```

## 4. QuPath and StarDist

`sg_stardist_parameter_map()` shows how every protocol parameter maps to the
QuPath StarDist extension, including what cannot be honoured:

```{r}
map <- sg_stardist_parameter_map("stardist.2d.v1", "qupath",
                                 normalize_scope = "tile",
                                 normalize_low = 1, normalize_high = 99)
map[, c("parameter", "target_option", "status")]
```

A QuPath program contains `run.json`, the Groovy templates and hashed
resources. segmantR never starts QuPath itself.

```{r}
calibrated <- img
calibrated$resolution <- list(x_um = 0.5, y_um = 0.5)
prog <- sg_export_qupath(run$mask, file.path(tempdir(), "qupath-import"),
                         image = calibrated, image_name = "nuclei.tif",
                         save_policy = "project", overwrite = TRUE)
prog$commands
```

For StarDist in QuPath, pass a model file (`.pb`, SavedModel or bioimage.io)
and an image describing the binding, then run the template headless or from
the Script Editor and import the results:

```{r, eval = FALSE}
sd <- sg_export_qupath("stardist.2d.v1", "qupath-stardist",
                       image = calibrated, image_name = "nuclei.tif",
                       model = "dsb2018_heavy_augment.pb",
                       normalize_scope = "tile", normalize_low = 1,
                       normalize_high = 99, pixel_size_um = 0.5,
                       save_policy = "project")
# In a shell, from qupath-stardist/:
#   QuPath script --project=project.qpproj --image="nuclei.tif" \
#     --args=run.json --save segmantR_stardist.groovy
back <- sg_import_qupath("qupath-stardist/out", image = calibrated,
                         expected_run_id = sd$run$run_id)
back$qupath$agreement
```

## 5. Capabilities

```{r}
caps <- sg_interchange_capabilities()
vapply(caps$profiles, function(p) p$status, character(1))
```

See `system.file("interop", "INTEROP.md", package = "segmantR")` for the
full contract, migration notes and security boundaries.
